Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-09 paper-new/towards-relational-model.tex
rename some files, progress on the relational model motivation section

\eric{Review this section}

In the previous section, we gave a denotational call-by-push-value model for the
terms of the gradually-typed lambda calculus using the tools of SGDT. In this
section, we build on the ideas used in defining a term model with the goal of
defining a model of the full gradually-typed lambda calculus, i.e., one that has
constructs that model type and term precision.

Recall that the cast calculus introduced in Section \ref{sec:GTLC} has not only
terms but also an \emph{inequational theory} with axioms for type precision and
term precision that constitute an abstract formulation of the graduality
property. Thus, our goal now is to build a denotational model interpreting these
axioms.
%

\subsection{Modeling Type and Term Precision}

% We will model type precision $c : A \ltdyn A'$ as a relation between the sets
% $\sem{A}$ and $\sem{A'}$.

To model term precision, we begin by equipping the denotation of every type with
an ordering relation. Since term precision is reflexive and transitive, and
since we identify terms that are equi-precise, we model value types as
partially-ordered sets and values $\Gamma \vdash V : A$ as \emph{monotone}
functions. Analogously, every error domain is now equipped with a partial
ordering for which the error element is the bottom element, and the map
$\theta_B \laterhs B \to B$ is now required to be monotone. 
%
\footnote{This requires us to define a notion of ``later'' for partially-ordered
sets $X$. The elements are simply ``later''-elements of $X$, and the ordering is
the obvious one, i.e., two later-elements are related if after one time step,
the elements they step to are related.}
%
Morphisms of error domains are morphisms of the underlying partially-ordered
sets that preserve the error element and $\theta$ map, as was the case in the
previous section.

@@

\begin{figure*}
    \begin{align*}
        &\eta\, x \semltbad \eta\, y \text{ if } 
            x \semlt y \\
        %		
        &\mho \semltbad l \\
        %
        &\theta\, \tilde{l} \semltbad \theta\, \tilde{l'} \text{ if } 
            \later_t (\tilde{l}_t \semltbad \tilde{l'}_t) \\
        %	
        &\theta\, \tilde{l} \semltbad \mho \text{ if } 
            \theta\, \tilde{l} = \delta^n(\mho) \text { for some $n$ } \\
        %	
        &\theta\, \tilde{l} \semltbad \eta\, y \text{ if }
            (\theta\, \tilde{l} = \delta^n(\eta\, x))
        \text { for some $n$ and $x : A$ such that $x \le y$ } \\
        %
        &\theta\, \tilde{l} \semltbad \mho \text { if } 
            \theta\, \tilde{l} = \delta^n(\mho) \text { for some $n$ } \\
        %	
        &\eta\, x \semltbad \theta\, \tilde{l'} \text { if }
            (\theta\, \tilde{l'} = \delta^n (\eta\, y))
        \text { for some $n$ and $y : A'$ such that $x \le y$ }
    \end{align*}
    \caption{The step-insensitive error ordering.}
    \label{fig:step-insensitive-error-ordering}
\end{figure*}

@@

\subsection{Problem: Failure of Transitivity}

The key issue with the above ordering is it is \emph{not} transitive, as we will
show below. We begin with a general ``no-go'' theorem about relations defined on
the guarded lift monad:

\begin{theorem}[No-go theorem for relations on $\li X$]\label{thm:no-go}
    Let $R$ be a relation on the guarded lift monad satisfying the following
    three properties:
    \begin{enumerate}
        \item Transitivity

        \item Congruence with respect to $\theta$: If $\later_t (\tilde{x}_t
        \binrel{R} \tilde{y}_t)$, then $\theta(\tilde{x}) \binrel{R}
        \theta(\tilde{y})$

        \item Closure under delay on both sides: If $x \binrel{R} y$, then $x
        \binrel{R} \delta y$ and $\delta x  \binrel{R} y$
    \end{enumerate}

    Then $R$ is trivial, i.e., $x \binrel{R} y$ for all $x : \li X$ and $y : \li
    Y$.

\end{theorem}

This theorem follows from two ``one-sided'' lemmas where we assume that the
relation $R$ is closed under delay on the left (resp. right) and show that
$(\fix\, \theta) \binrel{R} y$ for all $y$ (resp. $x \binrel{R} (\fix\, \theta)$
for all $x$). These lemmas in turn are proved by \lob-induction.

The theorem implies that the relation $\semltbad$ defined on $\li X \times \li
Y$ is \emph{not} transitive whenever the underlying relation $\semlt$ on $X$ and
$Y$ is not trivial, i.e., when there are $x_0$ and $y_0$ that are not related in
the relation $\semlt$ on $X$ and $Y$. To show this, first observe that
$\semltbad$ satisfies properties (2) and (3) in the above theorem. Congruence
with respect to $\theta$ is immediate from the definition, while closure under
delay can be proved using \lob-induction.
% TODO the proof of this uses tick irrelevance
Then since $x_0$ is not related to $y_0$, we have that $\eta\, x_0$ is not
related to $\eta\, y_0$ in $\semltbad$, i.e., $\semltbad$ is non-trivial. But if
it were transitive, then by the theorem, it would be trivial, a contradiction.
Thus, we are able to refute transitivity of $\semltbad$ provided the underlying
relation is non-trivial.

\subsection{Solution: Splitting into Two Relations}\label{sec:lock-step-and-weak-bisim}

>>>>

Now that we have seen how to model the terms without regards to
graduality, we next turn to how to enhance this model to provide a
compositional semantics that additionally satisfies graduality. By a
compositional semantics, we mean that we want to provide a
compositional interpretation of \emph{type} and \emph{term precision}
such that we can extract the graduality relation from our model of
term precision.

\subsection{A first attempt: Modeling Term Precision with Posets}

In prior work, New and Licata model the term precision relation by
equipping $\omega$-CPOs with a second ordering $\ltdyn$ they call the
``error ordering'' such that a term precision relation $M \ltdyn M' :
r(A)$ is modeled by $\sem{M} \ltdyn \sem{M'}$ in the semantics. Then
the error ordering on the free error domain $\li A$ is based on the
graduality property: essentially either $\sem{M}$ errors or both terms
diverge or both terms terminate with values related by the ordering
relation on $A$.
%
The obvious first try at a guarded semantics that models graduality
would be to similarly enhance our value types and computation types to
additionally carry a poset structure. For this value types can simply
be posets. For computation types, we want our error domains to be
posets as well. To consider what additional properties we want the
error domains to have, let's consider the particular case of the free
error domain $\li A$.

The ordering on $\li A$ should model the graduality property, but how
to interpret this is not so straightforward. The graduality property
as stated mentions divergence, but we only work indirectly with
diverging computations in the form of the abstract steps/think
maps. The closest direct encoding of the graduality property is an
ordering we call the \emph{step-insensitive error ordering}, defined
by guarded recursion in
Figure~\ref{fig:step-insensitive-error-ordering}.
%
Firstly, to model graduality, $\mho$ is the least element. Next, two
returned values are related just when they are in the ordering $\leq$
on $A$.
%
Two thinking elements are related just when they are related later.
%
The interesting cases are then those where one side is thinking and
the other has completed to either an error or a value.
%
Since the graduality property is \emph{extensional}, i.e., oblivious
to the number of steps taken, it is sensible to say that if one side
has terminated and the other side is thinking, that we must require
the thinking side to eventually terminate with a related behavior,
which is the content of the final $4$ cases of the definition.

\begin{figure*}
    \begin{align*}
        &\mho \semltbad l \text{ iff } \top \\
        %
        &\eta\, x \semltbad \eta\, y \text{ if } 
            x \semlt y \\
        %		
        &\theta\, \tilde{l} \semltbad \theta\, \tilde{l'} \text{ iff } 
            \later_t (\tilde{l}_t \semltbad \tilde{l'}_t) \\
        %	
        &\theta\, \tilde{l} \semltbad \mho \text{ iff } \exists n. \theta\, \tilde{l} = \delta^n(\mho) \\
        %	
        &\theta\, \tilde{l} \semltbad \eta\, y \text{ iff } \exists n. \exists x \leq y.
            (\theta\, \tilde{l} = \delta^n(\eta\, x)) \\
        %
        &\theta\, \tilde{l} \semltbad \mho \text { iff } 
            \exists n. \theta\, \tilde{l} = \delta^n(\mho) \\
        %	
        &\eta\, x \semltbad \theta\, \tilde{l'} \text { iff }
            \exists n. \exists y \geq x. (\theta\, \tilde{l'} = \delta^n (\eta\, y))
    \end{align*}
    \caption{The step-insensitive error ordering.}
    \label{fig:step-insensitive-error-ordering}
\end{figure*}

This definition of $\semltbad$ is an entirely reasonable attempt to
model graduality: for instance it is true that if $l \semltbad l'$ for
$\li \mathbb{N}$ where $\mathbb{N}$ is the flat poset whose ordering
is equality, then the bigstep semantics we derive for $l, l'$ do in
fact satisfy the intended graduality relation.
%
However, there is a major issue with this definition: it's not a poset
at all: it is not anti-symmetric, but more importantly it is
\emph{not} transitive!
%
To see why, we observe the following undesirable property of any
relation that is transitive and ``step-insensitive'' on one side:
\begin{theorem}\label{thm:no-go}
  Let $R$ be a binary relation on the free error domain $U(\li A)$. If
  $R$ satisfies the following properties
  \begin{enumerate}
  \item Transitivity
  \item $\theta$-congruence: If $\later_t (\tilde{x}_t \binrel{R} \tilde{y}_t)$, then $\theta(\tilde{x}) \binrel{R} \theta(\tilde{y})$.
  \item Right step-insensitivity: If $x \binrel{R} y$ then $x \binrel{R} \delta y$.
  \end{enumerate}
  Then for any $l : U(\li A)$, we have $l R \Omega$. If $R$ is left
  step-insensitive instead then $\Omega \binrel{R} x$.
\end{theorem}
\begin{proof}
  By L\"ob induction, in the appendix.
\end{proof}
\begin{corollary}
  Let $R$ be a binary relation on $U(\li A)$. If $R$ satisfies
  transitivity, $\theta$-congruence and left and right
  step-insensitivity, then $R$ is the total relation: $\forall x, y. x
  \binrel{R} y$.
\end{corollary}
\begin{proof}
  $x \binrel{R} \Omega$ and $\Omega \binrel{R} y$ by the previous
  theorem. Then by transitivity $x \binrel R y$.
\end{proof}

This in turn implies $\semltbad$ is not transitive, since it is a
$\theta$-congruence and step-insensitive on both sides, but not
generally trivial: e.g., for the flat poset on $\mathbb N$, $\eta 0 \semltdbad
\eta 1$ is false.
%
This shows definitively that we cannot provide a compositional
semantic graduality proof based on types as posets if we use this
step-insensitive ordering on $\li A$.

\subsection{Why transitivity is essential}

At this point we might try to weaken from our initial attempt at
providing a poset semantics to a semantics where types are equipped
with merely a reflexive relation, and continue using the
step-insensitive ordering. However some level of transitivity is
absolutely essential to providing compositional reasoning, and is used
pervasively in the New-Licata development.
%
The reason that it is essential has to do with the details of the
graduality proof, so first we expand on how we prove graduality
compositionally. First of all, we need to extend our compositional
semantics of types and terms to also a compositional semantics of
\emph{type precision derivations} and \emph{term precision}.
%
Sticking to a poset-based semantics, if our value types are posets, we
then want our precision derivations $c : A \ltdyn A'$ to be
\emph{poset relations} on $A,A'$, that is relations between the
underlying sets of $A,A'$ that are \emph{downward-closed} in $A$ and
\emph{upper-closed} in $A'$. The obvious example of such a relation is
the poset ordering itself $\leq_A : A \rel A'$...\max{TODO: Eric pick
  up from here. Go into the squares for casts and how representability
  gives us a compositional approach to proving them}

\subsection{Resolution: Splitting Error Ordering and Bisimilarity}\label{sec:lock-step-and-weak-bisim}

\subsection{Modeling Type and Term Precision}
\max{TODO: figure out where this stuff goes}
% We will model type precision $c : A \ltdyn A'$ as a relation between the sets
% $\sem{A}$ and $\sem{A'}$.

To model term precision, we begin by equipping the denotation of every type with
an ordering relation. Since term precision is reflexive and transitive, and
since we identify terms that are equi-precise, we model value types as
partially-ordered sets and values $\Gamma \vdash V : A$ as \emph{monotone}
functions. Analogously, every error domain is now equipped with a partial
ordering for which the error element is the bottom element, and the map
$\theta_B : \laterhs B \to B$ is now required to be monotone. 
%
Morphisms of error domains are morphisms of the underlying partially-ordered
sets that preserve the error element and $\theta$ map, as was the case in the
previous section.

@@

@@
